Generalized spaces for constructive algebra

### 2020-04-08 paper-herrsching2019/main.tex
Anonymous Noetherianity

\begin{exercise}{A geometric interpretation of double negation}%
Let~$\varphi$ be a formula over an open~$U$ of a locale~$X$.
\begin{alphlist}[(c)]
\item Show that there is a largest open,
denoted~``$\llbracket\varphi\rrbracket$'', such that~$U \models \varphi$.\smallskip

@@

\paragraph{Double negation stability.}
XXX explain that it is surprising and explain usefulness, then give proof

\begin{proposition}The sheaf of rings~$A^\sim$ has $\neg\neg$-stable equality:
\[ \Spec(A) \models \forall x,y\?A^\sim\_ \neg\neg(x = y) \Rightarrow x = y. \]
\end{proposition}

\begin{proof}We argue internally. Let~$x,y \? A^\sim$ be given such
that~$\neg\neg(x = y)$. We will verify that~$x - y$ is not invertible; by the
field condition, this will imply~$x - y = 0$, hence~$x = y$.

So assume that~$x - y$ is invertible. Then~$\neg(x - y = 0)$, since if~$x - y =
0$, then~$1 = (x-y) (x-y)^{-1} = 0$ in contradiction to~$1 \neq 0$. This is a
contradiction to the assumption~$\neg\neg(x-y=0)$.\end{proof}

@@

% Ex.: Z finitely generated

This definition is at odds with the idea that constructive mathematics should
be informative, because it only expresses that it is impossible for no finite
generating family to exist without requiring that such a family should be given
-- it may remain \emph{anonymous}.\footnote{We borrowed the term ``anonymous''
from type theory, where it is used with a similar meaning (see for instance
Ref.~\refcite{kraus-escardo-coquand-altenkirch:anonymous}). However, there is a
subtle difference: The conjunction of ``there is at most one element~$x$'' and
``there is anonymously an element~$x$ in the sense of type theory'' implies
that there actually exists an element~$x$. In contrast, \notnot existence does
not.}

However, it is still useful in constructive algebra when interpreted by the
sheaf semantics: If~$A$ is reduced, then~$A^\sim$ is anonymously Noetherian.
XXX

%\begin{ex}There is an intuitionistic proof that the ring~$\ZZ$ is anonymously
%Noetherian: Let~$\aaa \subseteq \ZZ$ be any ideal. Under the assumption that
%either there exists a nonzero element in~$\aaa$ or not, the ideal~$\aaa$ is
%\notnot finitely generated, even \notnot principal: For in the first case, a
%minimal element~$d$ of~$\aaa \cap \NN^+$ (which \notnot exists) witnesses~$\aaa
%= (d)$. In the second case the ideal~$\aaa$ is the zero ideal. Since the
%assumption is \notnot satisfied, the ideal~$\aaa$ is \notnot \notnot finitely
%generated, so \notnot finitely generated. (We remark on this proof scheme on
%page~\pageref{proof-scheme-boxed-statements}.) \end{ex}
%
%It appears that this Noetherian condition has not been studied in the
%literature on constructive algebra. Indeed, from the philosophical and the
%computational point of view on constructive mathematics, the notion of an anonymously
%Noetherian ring is not very useful: From an intuitionistic proof that a given
%ideal is finitely generated one can mechanically extract explicit generators,
%thereby satisfying computational or philosophical demands (``the generators are
%really there''). In contrast, an intuitionistic proof that a given ideal is
%\notnot finitely generated doesn't contain computational content in general.
%
%Along the same lines, one could dismiss the fact (proved below) that Hilbert's
%basis theorem, stating that~$A[X]$ is Noetherian if~$A$ is, holds for the
%notion of anonymously Noetherian rings. In fact, one could feel mocked by this
%version of Hilbert's basis theorem: It promises that any ideal of~$A[X]$
%``has'' finitely many generators in some Platonic sense (without providing any
%clue on how one might go on to find the generators -- they remain \emph{anonymous}), provided
%that any ideal of~$A$ ``has'' finitely many generators in the same
%sense.\footnote{}

\begin{proposition}The sheaf of rings~$A^\sim$ is anonymously
Noetherian.\end{proposition}

\begin{proof}We argue internally. Let~$\aaa \subseteq A^\sim$ be an ideal.
Assume that~$\aaa$ is not finitely generated. We will verify that then~$\aaa =
(0)$, hence that~$\aaa$ is finitely generated; contradiction.

@@

@@

This increase in complexity should be properly appreciated: The sheaf semantics
allows us to harness convoluted properties in simple language. The unrolled
monstrosity in the proof of Proposition~\ref{prop:field-property} is not
memorable at all and unlikely to find its way into human-written proofs.
However, in the form~``$A^\sim$ is a field'', it becomes easily accessible.

This state of affairs is even more pronounced for higher-order formulas such as
the condition to be anonymously Noetherian. For instance, XXX

\begin{quote}
For any element~$f \in A$ and any (not necessarily quasicoherent) sheaf of
ideals~$\J \hookrightarrow A^\sim[X_1,\ldots,X_n]|_{D(f)}$: If
\begin{quote}
for any element~$g \in A$ the condition that
\begin{quote}
the sheaf~$\J$ is of finite type on~$D(fg)$
\end{quote}
implies that~$fg$ is nilpotent,
\end{quote}
then~$f$ is nilpotent.
\end{quote}

>>>>

\begin{exercise}{A geometric interpretation of double negation}%
\label{ex:negneg-interpretation}%
Let~$\varphi$ be a formula over an open~$U$ of a locale~$X$.
\begin{alphlist}[(c)]
\item Show that there is a largest open,
denoted~``$\llbracket\varphi\rrbracket$'', such that~$U \models \varphi$.\smallskip

@@

\paragraph{Double negation stability.} Generally in constructive mathematics,
two elements being \notnot equal does not imply that they are actually equal:
\begin{enumerate}
\item Sheaves give counterexamples: The interpretation of the statement that
two given sections of a sheaf are \notnot equal with the sheaf semantics is
that they agree on a dense open (Exercise~\ref{ex:negneg-interpretation}). This
does not generally imply that they agree on all of their domain.
\item Realizability gives counterexamples: Realizers of negated statements are
always uninformative; in the case that a statement~$\varphi$ has a realizer at
all, the statement~$\neg\neg\varphi$ is realized by any number whatsoever.
Only in certain cases, where a realizer can be algorithmically reconstructed
from the mere promise that a realizer exists, is~$\neg\neg\varphi \Rightarrow
\varphi$ realizable.
\end{enumerate}

Hence statement~(\ref{item:negneg-stable}) comes at a surprise. It is useful to
import some of classical logic into the constructive setting of sheaves
on~$\Spec(A)$. For instance, the law of excluded middle is constructively valid
in the form
\[ \neg\neg(\varphi \vee \neg\varphi). \]
Combined with the observation
\[ \bigl(\neg\neg\chi \wedge (\chi \Rightarrow \neg\neg\psi)\bigr) \Longrightarrow \neg\neg\psi \]
we can hence freely use the law of excluded middle in constructive arguments --
but at the price that the conclusion~$\neg\neg\psi$ is doubly negated, which is
why typically this observation is not of much use.

However, double negation stability of equality does allow us to conclude~$\psi$
from the a~priori much weaker premiss~$\neg\neg\psi$, if~$\psi$ is of the
form~$(s =_{A^\sim} t)$.

This stability is a consequence of the field property. In case that the
ring~$A$ is not reduced and hence~$A^\sim$ not a field, a substitute is given
in Exercise~\ref{ex:local-global}. We stress that the proof of
statement~(\ref{item:negneg-stable}) does not require double negation stability
of~$A$ on the metalevel; in fact, it is fully constructive.

\begin{proposition}If~$A$ is reduced, then the sheaf of rings~$A^\sim$ has $\neg\neg$-stable equality:
\[ \Spec(A) \models \forall x,y\?A^\sim\_ \neg\neg(x = y) \Rightarrow x = y. \]
\end{proposition}

\begin{proof}We argue internally in the universe of sheaves, using the sheaf
semantics. Let~$x,y \? A^\sim$ be given such
that~$\neg\neg(x = y)$. We will verify that~$x - y$ is not invertible; by the
field condition, this will imply~$x - y = 0$, hence~$x = y$.

So assume that~$x - y$ is invertible. Then~$\neg(x - y = 0)$, since if~$x - y =
0$, then~$1 = (x-y) (x-y)^{-1} = 0$ in contradiction to~$1 \neq 0$. Hence we have a
contradiction to the assumption~$\neg\neg(x-y=0)$.\end{proof}

@@

\begin{example}The assumption that any ideal of~$\ZZ$ is finitely generated implies
the law of excluded middle. But we can prove, constructively, that~$\ZZ$ is
anonymously Noetherian: Let~$\aaa$ be an ideal of~$\ZZ$ and assume that~$\aaa$
is not finitely generated. Then~$1 \not\in \aaa$, as else~$\aaa = (1)$ would be
finitely generated. Also~$2 \not\in \aaa$, as else (noting
that~$1\not\in\aaa$),~$\aaa = (2)$. Continuing in this manner, we may
conclude~$n \not\in \aaa$ for every number~$n \geq 1$. Hence~$\aaa = (0)$. But
this is a contradiction to the assumption that~$\aaa$ is not finitely
generated.
\end{example}

Definition~\ref{defn:anonymously-noetherian} is at odds with the idea that constructive mathematics should
be informative: It only expresses that it is impossible for no finite
generating family to exist without requiring that such a family should be given
-- it may remain \emph{anonymous}.\footnote{We borrowed the term ``anonymous''
from type theory, where it is used with a similar meaning (see for instance
Ref.~\refcite{kraus-escardo-coquand-altenkirch:anonymous}). However, there is a
subtle difference: The conjunction of ``there is at most one element~$x$'' and
``there is anonymously an element~$x$ in the sense of type theory'' implies
that there actually exists an element~$x$. In contrast, \notnot existence does
not.} And while Hilbert's basis theorem, stating that if a ring~$B$ is
Noetherian then so is the polynomial ring~$B[X]$, does have a constructive
proof for the anonymous version of the Noetherian condition,\footnote{For
instance, a careful reading of the textbook proof given
in~\cite[Theorem~7.5]{atiyah-macdonald:commutative-algebra} shows that the
basis theorem holds intuitionistically as stated.} this result seems empty: For
it just transforms one unconcrete promise (generators of each ideal of~$B$
exist somewhere, platonically) into another (generators for ideals of~$B[X]$
exist somewhere, platonically).

Curiously, the anonymous version of the Noetherian condition is still useful in
constructive algebra when interpreted by the sheaf semantics: This is firstly
because it just so happens that~$A^\sim$ is anonymously Noetherian if~$A$ is reduced,
and secondly because the interpretation of double negation with the sheaf
semantics does have nontrivial content (as evidenced by
statement~(\ref{item:negneg-stable})).

\begin{proposition}\label{prop:anon-noeth}
If~$A$ is reduced, then the sheaf of rings~$A^\sim$ is anonymously
Noetherian.\end{proposition}

\begin{proof}We argue internally, under the sheaf semantics. Let~$\aaa \subseteq A^\sim$ be an ideal.
Assume that~$\aaa$ is not finitely generated. We will verify that then~$\aaa =
(0)$, hence that~$\aaa$ is finitely generated; contradiction.

@@

\begin{remark}The condition for a ring to be (anonymously) Noetherian is a
higher-order condition and hence out of scope of the first-order version of the
sheaf semantics presented in Section~\ref{sect:sheaf-semantics}. However, the
general version of the sheaf semantics is sound with respect to intuitionistic
higher-order logic, hence the proof of Proposition~\ref{prop:anon-noeth} is
valid.
\end{remark}

@@

This increase in complexity should be properly appreciated: The sheaf semantics
allows us to harness convoluted properties in simple language. The unrolled
monstrosity in the proof of Proposition~\ref{prop:field-property} is not
memorable at all and unlikely to find its way into human-written proofs.
However, in the form~``$A^\sim$ is a field'', it becomes easily accessible.
This state of affairs is the more pronounced the more negations and especially
double negations occur in the proof.

%\begin{quote}
%For any element~$f \in A$ and any (not necessarily quasicoherent) sheaf of
%ideals~$\J \hookrightarrow A^\sim[X_1,\ldots,X_n]|_{D(f)}$: If
%\begin{quote}
%for any element~$g \in A$ the condition that
%\begin{quote}
%the sheaf~$\J$ is of finite type on~$D(fg)$
%\end{quote}
%implies that~$fg$ is nilpotent,
%\end{quote}
%then~$f$ is nilpotent.
%\end{quote}

\begin{example}For future reference, and also for giving a different
perspective on statement~(\ref{item:negneg-stable}), we unroll here the
interpretation of~$f \models \neg\neg\varphi$. It is:
\begin{quote}
For any element~$g \in A$, \\
${\qquad}$ if for any element~$h \in A$, \\
${\qquad\qquad}$ if~$fgh \models \varphi$, \\
${\qquad\qquad}$ then~$fgh$ is nilpotent, \\
${\qquad}$ then~$fg$ is nilpotent.
\end{quote}
In the case that~$A$ is reduced and~$f$ is the unit element of~$A$, this
condition implies (by setting~$g \defeq 1$): ``If~$h = 0$ is the only element of~$A$
such that~$h \models \varphi$, then~$1 = 0$ in~$A$.''
\end{example}
